Solution of thermodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2014-05-08 pituba/article/article_spe.tex
america 2

% Review
\documentclass[authoryear,preprint,review,12pt]{elsarticle}
%\documentclass[final,authoryear,5p,times,twocolumn,11pt]{elsarticle}

@@

\begin{abstract}
This work presents a derivation of the classical material balance equation starting from black-oil equations in integral form. The existing link between both approaches is examined in detail, including the necessary step of averaging PVT properties. An interpretation for pressure in material balance calculations is provided and the choice of a reference depth (datum) is discussed. The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented. A scheme for numerical solution is also proposed.
\end{abstract}

@@

Black-oil equations arise from the generalization of the zero dimensional concept to a three-dimensional model that incorporates Darcy's law and entails the same approach for phase equilibrium \citep{aziz1979petroleum}.

@@

Although many applications of material balance equation have been reported in the literature, few explore its connection with black-oil equations. \cite{ertekin2001basic} briefly addressed this issue arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \mathbf{g}$) were neglected and if PVT properties were averaged.

@@

We employ the same approach of varying bubble-point pressure widely used in reservoir simulation that is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}. The employed approach extends material balance equation to saturated reservoirs under pressure maintenance (e.g. fluid injection). The applicability, however, depend on how important non-equilibrium effects are to the phase equilibrium.

@@

%\section{Black-Oil Equations}
%
%Here it is used a thermodynamic fluid model known as Black-Oil \cite{blackoil}. It consists of a simplified treatment of hydrocarbon composition and phase equilibrium. Hydrocarbon species are grouped into two pseudo-components called oil and gas, being a pseudo-component one which consists of more than one chemical molecule.
%
%Oil is composed by hydrocarbon species that remain in liquid phase for standard surface pressure and temperature. The other hydrocarbon species in gaseous phase for the same surface conditions are called gas.
%
%The mixture of reservoir fluid is then composed of oil, gas and water components and can present three fluid phases: oil, gas and water. Oil phase may be composed of oil and gas components, gaseous phase may be composed uniquely by gas component and water phase uniquely by water component.
%
%In addition to the three fluid components and phases, there is a solid phase composed by the rock.
%
%The fluid phases may be defined in a more general framework so as to account for dissolution of gas in water or water vapour, for instance. Such possibilities are not dealt with in this work.
%
%For some volume containing one of the three fluid phases, the saturation of the fluid phase $k$ is defined as
%\begin{equation}\label{eq: saturation}
%S_{k}=\frac{\phi_k}{\sum_{i \neq solid} \phi_i} \, ,
%\end{equation}
%where $\phi_i$ is the volume fraction of the fluid phase $i$.
%
%In order to apply the volume balance, a control volume ($\Omega$) is defined containing all the hydrocarbon zone. It consists of the region where hydrocarbon saturation is greater than zero for some arbitrarily small volume. This reservoir definition excludes the pure water zone, which is referred to as the aquifer.
%
%
%Right hand side of Equation \eqref{eq: matbal1}, \textit{rock and fluid expansion}, is a function of phase mass transfers and compressibilities. Both should be obtained from a thermodynamic model accounting for mass transfer among phases and the relation between phase volumes and pressure. Temperature is seldom considered to vary during reservoir production and it is held constant and equal in all phases, not being an independent property of the control volume state.
%
%Such thermodynamic model is presented in detail in \cite{blackoil} and it is commonly referred to as Black-Oil Equations. The main assumption behind them is that phase and pressure equilibria are instantaneously achieved inside the control volume, i.e., mass transfer among phases reaches equilibrium and internal forces equilibrate external ones fast enough in comparison to the interval of time between two observations of control volume properties.
%
%Pressure in each phase may differ if capillary forces are taken into account, for what extra constitutive equations relating each phase pressures would be needed.
%Capillary pressure would be function of porous medium properties given by some porous medium model, e.g., permeability and porosity; phase properties, e.g., surface tension and wettability, and the thermodynamic state.
%
%Capillary pressure is very often assumed to be function of phase saturations only and, hence, the thermodynamic state of the control volume would be completely determined by the pressure of one phase (e.g. the wetting phase) and phase saturations.
%
%Black-Oil Equations do not assume water and oil phases becoming miscible for any thermodynamic state.
%
%Capillary pressure dependence on phase saturation only is often considered as a weak assumption in contradiction with experimental observations. The hysteresis of capillary pressure exhibited when either imbibition or drainage occurs is not theoretically explained and it is accounted for by empirical relations or it is rather simply neglected.
%
%\cite{hilfer1} has proposed the inclusion of an energy equation.

@@

\subsection{Black-Oil equations}

@@

\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV + \oint_{\partial \Omega} \rho_w \left( \mathbf{u}_w \cdot \mathbf{dS} \right) = \rho_w^{ST}Q_w^{ST} \, ;
\end{equation}

\item oil component:
\begin{equation}\label{eq: So1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV + \oint_{\partial \Omega} \rho_o Y_o \left(\mathbf{u}_o \cdot \mathbf{dS} \right)= \rho_o^{ST}Q_o^{ST} \, ;
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV + \oint_{\partial \Omega} \left( \rho_o Y_g \mathbf{u}_o + \rho_g \mathbf{u}_g\right) \cdot \mathbf{dS}= \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}
Mass concentration of oil and gas components in oil phases are denoted as $Y_o$ and $Y_g$, respectively.

@@

A common simplification in black-oil formulations is to restrict mass transport to macroscopic advection. No diffusive or dispersive term is included in the mass balance equations. This simplification must be looked with care when gradients of species concentration are important. For instance, when gas is being dissolved in oil phase in a reservoir under repressurization, this condition is not generally true.

@@

%Mass balance equations for each fluid phase are \citep{ertekin2001basic}:
%% Oleo
%\begin{equation}\label{eq: So1}
%\frac{\partial}{\partial t}\left(\frac{\phi(p_P) S_o}{B_o(p_o,P_b)}\right)+\nabla\cdot\left(\frac{\mathbf{u}_o}{B_o (p_o,P_b)}\right)=q_{os} \, ,
%\end{equation}
%% Gas
%\begin{equation}
%\begin{split}\label{eq: Sg1}
%\frac{\partial}{\partial t}\left[\phi(p_P)\left(\frac{S_g}{B_g (p_g)} +\frac{S_o R_s (p_o,P_b)}{B_o (p_o,P_b)}\right)\right] \\
%+\nabla\cdot\left(\frac{\mathbf{u}_o R_s (p_o,P_b)}{B_o (p_o,P_b)}+\frac{\mathbf{u}_g}{B_g (p_g)}\right)=q_{gs} \, ,
%\end{split}
%\end{equation}
%% Agua
%\begin{equation}\label{eq: Sw1}
%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)+\nabla\cdot\left(\frac{\mathbf{u}_w}{B_w(p_w)}\right)=q_{ws} \, ,
%\end{equation}
%
%where saturations must satisfy the constraint
%\begin{equation}\label{eq: S11}
%S_o+S_w+S_g=1 \, .
%\end{equation}
%
%Bubble-point pressure ($P_b$) is here defined as the minimum pressure for which gas phase is locally absent, i.e., $S_g=0$. Its local value is implicitly given by the solubility ratio ($R_s$) satisfying the equation:
%\begin{equation}\label{eq: Pb}
%R_s(p_o=P_b,P_b) = \frac{B_g(p_g)}{B_o(p_o,P_b)}\frac{S_g}{S_o} +R_s(p_o,P_b) \, .
%\end{equation}
%It must be remarked that bubble-point pressure is sometimes defined in a different manner as the bubble-point pressure for the hydrocarbon composition in oil phase only. According to this latter definition, $P_b = p_o$ for any saturated reservoir.
%
%Black-oil equations make the assumption of instantaneous mass transfer between oil and gas phases, what is only reasonable for a reservoir under depletion, when oil and gas components are homonegeously distributed in oil phase. Once constant temperature and fluid composition are assumed, diffusion and advection of species and energy become unimportante for phase equilibrium calculations.
%
%If gradients of species concentration are present within oil phase, however, as it occurs in saturated reservoirs under pressure maintenance, a combined mechanism of advection and diffusion exists and those are absent of black-oil equations. Some heuristic models have been proposed for problems where this issue is critical, as for gas injection projects. \cite{utseth1981numerical} has proposed to limit $dR_s/dt$ and neglect transport of gas component in oil phase for a simplified treatment.

Momentum balance for each phase is given by an extension of Darcy's law to multiphase flow. This extension consists in multiplying absolute (or single-phase) permeability ($\mathbf{k}$) by a scalar function called phase relative permeability ($k_{r,\alpha}$):

\begin{equation}\label{eq: exdarcy}
\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right) \, , \quad \alpha=o,w,g \, .
\end{equation}

@@

Temperature is thus a constant prescribed value and no energy balance equation is needed. Also, it follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if we restrict component number to the two pseudo-components. As temperature is assumed constant, oil phase density in saturated states is a function of pressure or composition only. The same is valid for mass concentrations of gas and oil components in oil phase.

@@

In order to close the system of equations, a connection among different phase pressures must be provided. Such expressions are referred to as capillary pressures and are usually employed in the following forms \citep{aziz1979petroleum}:
%capillary
\begin{equation}\label{eq: capillary_ow}
p_{c,ow}\left(\mathbf{k},\phi,S_w\right)=p_o-p_w \, ,
\end{equation}
\begin{equation}\label{eq: capillary_og}
p_{c,go}\left(\mathbf{k},\phi,S_g\right)=p_g-p_o \, .
\end{equation}

@@

%It is well known that black-oil equations possess discontinuous solutions in saturation, bubble pressure and solubility ratio fields. Discontinuities (or shocks) are introduced by initial or boundary conditions and are propagated throughout flow domain due to the hyperbolicity of equations.
%
%In initial conditions, if hydrostatic equilibrium is imposed and no capillary pressure is included, that is,
%\begin{equation}
%p_{c,ow}=p_{c,og}=0 \quad \longrightarrow \quad p_o=p_w=p_g=p\, ,
%\end{equation}
%saturation discontinuities will be present due to gravity segregation.
%
%When capillary pressure is present and it is function of phase saturations, a diffusive term is introduced in mass balance equations and shocks originated by phase displacement become smoothed:
%\begin{equation}
%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)-\nabla\cdot\left[\frac{k_{r,w} \mathbf{k}}{B_w(p_w)\mu_w} \left(\nabla p_o - \rho_w g\nabla z -\underbrace{\frac{d p_c}{d S_w}\nabla S_w}_{\text{water diffusion}}\right)\right]=q_{ws} \, .
%\end{equation}
%
%Discontinuities in saturation may still occur, however, if capillary pressures are functions of discontinuous porous medium properties ($phi$ and $\mathbf{k}$) or if phase change takes place, as diffusion of gas dissolved in oil phase is not allowed.
%Cite Bear and Herrera.

@@

We start assuming mechanical equilibrium, i.e., no bulk motion:
\begin{equation}
\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right)\equiv 0 \, .
\end{equation}
This condition implies that pressure gradients must equal gravity force per unit volume:
\begin{equation}\label{eq: p=rhogh}
\nabla p_\alpha = \rho_\alpha g\nabla z
\end{equation}
and, consequently, pressure is a function of depth only:
\begin{equation}\label{eq: p=rhogh}
\int_{p_\alpha^0}^{p_\alpha} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'},T\right)} = g\left(z - z_0\right) \, .
\end{equation}

@@

\begin{equation}
\bar{\phi} (t) = \frac{1}{\Omega}\int_{\Omega}\phi \left(\mathbf{x},t\right) dV \, .
\end{equation}

We proceed omitting function arguments:
\begin{equation}
\bar{S_\alpha}\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}S_\alpha \phi dV \, ,
\end{equation}

\begin{equation}
\bar{\rho_\alpha}\bar{S_\alpha}\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha S_\alpha \phi dV \, ,
\end{equation}

\begin{equation}
\bar{\rho_\alpha}\bar{Y_{\alpha,k}}\bar{S_\alpha}\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha Y_{\alpha,k} S_\alpha \phi dV \, .
\end{equation}

@@

\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw3}
\frac{\partial}{\partial t} \left( \bar{\rho_w}\bar{S_w}\bar{\phi} \Omega \right) = \rho_w^{ST}Q_w^{ST}
\end{equation}

\item oil component:
\begin{equation}\label{eq: So3}
\frac{\partial}{\partial t} \left( \bar{\rho_o}\bar{Y_o}\bar{S_o}\bar{\phi} \Omega \right) = \rho_o^{ST}Q_o^{ST}
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg3}
\frac{\partial}{\partial t} \left[\left( \bar{\rho_o}\bar{Y_g} \bar{S_o} + \bar{\rho_g} \bar{S_g} \right) \bar{\phi} \Omega\right] = \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}

@@

\begin{equation}\label{eq: Bw}
B_w = \frac{\rho_w^{ST}}{\bar{\rho_w}} \, ,
\end{equation}

\begin{equation}\label{eq: Bo}
B_o = \frac{\rho_o^{ST}}{\bar{Y_o} \bar{\rho_o}} \, ,
\end{equation}

\begin{equation}\label{eq: Bg}
B_g = \frac{\rho_g^{ST}}{\bar{\rho_g}} \, .
\end{equation}

@@

\begin{equation}
R_s = \frac{\rho_o^{ST} / \bar{Y_o}}{\rho_g^{ST} / \bar{Y_g}} \, ,
\end{equation}

@@

\begin{equation}\label{eq: Sw4}
\frac{\partial}{\partial t} \left(\frac{\bar{S_w} V_p}{B_w} \right) = Q_w^{ST} \, ,
\end{equation}

\begin{equation}\label{eq: So4}
\frac{\partial}{\partial t} \left(\frac{\bar{S_o} V_p}{B_o} \right) = Q_o^{ST} \, ,
\end{equation}

\begin{equation}\label{eq: Sg4}
\frac{\partial}{\partial t} \left[ V_p \left(\frac{\bar{S_g}}{B_g} + \frac{\bar{S_o} R_s}{B_o} \right) \right] = Q_g^{ST} \, ,
\end{equation}

\begin{equation}\label{eq: S14}
\bar{S_w}+\bar{S_o}+\bar{S_g}=1 \, .
\end{equation}

@@

For answering this question, we examine the two roles played by pressure in material balance equation: computation of averaged phase density and phase equilibrium. We shall see that for the purpose of computing the averaged phase density, we may define some reference depth and use the corresponding local density as an approximation to the averaged density without introducing any important error. Pressure increases with depth according to the hydrostatic gradient. Consequently, density also increases but by a negligible amount.

@@

For the case of an oil reservoir covered by a gas cap, the reference depth may be defined, with advantage, at the gas-oil contact. The point here is that at phase contact both gas and oil densities may be computed using the same pressure, i.e., the corresponding local value.

@@

From Eq.~\eqref{eq: dRsdz2}, we see that the variation of pressure with depth may not be negligible for the phase equilibrium,  specially for reservoirs of light oils with large depth ranges. In such cases, using an averaged value for the solubility ratio may be worth the accuracy gain in dissolved gas volume.

@@

\section{Solution of Material balance equation}
A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We shall start defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
\begin{equation}\label{eq: residual}
\varepsilon(p_2) = \frac{E+F_i+C-F_p}{V_{p1}} \, .
\end{equation}
We proceed showing that $\varepsilon$ is a monotonic function provided some conditions on fluid properties are observed and that, as a consequence, only one root of $\varepsilon$ exists at most. Equivalently, we may say that only one solution exists for material balance equation.

@@

\subsection{Mathematical Aspects of the Residual Function}

>>>>

% Review
\documentclass[authoryear,preprint,review,10pt]{elsarticle}
%\documentclass[final,authoryear,5p,times,twocolumn,11pt]{elsarticle}

@@

%
%This work derives the classical material balance equation from the more general black-oil model. It is shown that spatially averaged PVT properties are required for the material balance approach in general, but that for many cases the averaging is simply negligible for practical purposes.  An interpretation for pressure in material balance equation is provided and its relation to a reference depth is examined. The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented. A scheme for numerical solution is also proposed.

\begin{abstract}
This work derives the classical material balance equation from the more general black-oil model. The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented. It is shown that spatially averaged PVT properties are required for material balance approach in general, though the averaging process may be neglected for many practical purposes.  An interpretation for pressure in material balance equation is provided and its relation to a reference depth is examined. Finally, a scheme for numerical solution is proposed.

Material balance equation and black-oil models are approaches widely employed in reservoir engineering though few works explore their connection beyond the general intuitive notion. 
%
%This work derives the classical material balance equation from a black-oil model. We examine the existing link between both approaches in detail, including the necessary step of averaging PVT properties. An interpretation for pressure in material balance calculations is provided and the choice of a reference depth (datum) is discussed. The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented. A scheme for numerical solution is also proposed.
\end{abstract}

@@

%Black-oil equations arise from the generalization of the zero dimensional concept to a three-dimensional model that incorporates Darcy's law and entails the same approach for phase equilibrium \citep{aziz1979petroleum}.

@@

Although many applications of material balance equation have been reported in the literature, few explore its connection with black-oil equations. \cite{ertekin2001basic} briefly addressed this issue arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \vec{g}$) were neglected and if PVT properties were averaged.

@@

We employ the same approach of varying bubble-point pressure widely used in reservoir simulation that is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}. The employed approach extends material balance equation to saturated reservoirs under pressure maintenance (e.g. fluid injection). The applicability, however, depends on how important non-equilibrium effects are to the phase equilibrium.

@@

%\section{Black-oil equations}
%
%Here it is used a thermodynamic fluid model known as Black-Oil \cite{blackoil}. It consists of a simplified treatment of hydrocarbon composition and phase equilibrium. Hydrocarbon species are grouped into two pseudo-components called oil and gas, being a pseudo-component one which consists of more than one chemical molecule.
%
%Oil is composed by hydrocarbon species that remain in liquid phase for standard surface pressure and temperature. The other hydrocarbon species in gaseous phase for the same surface conditions are called gas.
%
%The mixture of reservoir fluid is then composed of oil, gas and water components and can present three fluid phases: oil, gas and water. Oil phase may be composed of oil and gas components, gaseous phase may be composed uniquely by gas component and water phase uniquely by water component.
%
%In addition to the three fluid components and phases, there is a solid phase composed by the rock.
%
%The fluid phases may be defined in a more general framework so as to account for dissolution of gas in water or water vapour, for instance. Such possibilities are not dealt with in this work.
%
%For some volume containing one of the three fluid phases, the saturation of the fluid phase $k$ is defined as
%\begin{equation}\label{eq: saturation}
%S_{k}=\frac{\phi_k}{\sum_{i \neq solid} \phi_i} \, ,
%\end{equation}
%where $\phi_i$ is the volume fraction of the fluid phase $i$.
%
%In order to apply the volume balance, a control volume ($\Omega$) is defined containing all the hydrocarbon zone. It consists of the region where hydrocarbon saturation is greater than zero for some arbitrarily small volume. This reservoir definition excludes the pure water zone, which is referred to as the aquifer.
%
%
%Right hand side of Equation \eqref{eq: matbal1}, \textit{rock and fluid expansion}, is a function of phase mass transfers and compressibilities. Both should be obtained from a thermodynamic model accounting for mass transfer among phases and the relation between phase volumes and pressure. Temperature is seldom considered to vary during reservoir production and it is held constant and equal in all phases, not being an independent property of the control volume state.
%
%Such thermodynamic model is presented in detail in \cite{blackoil} and it is commonly referred to as Black-Oil Equations. The main assumption behind them is that phase and pressure equilibria are instantaneously achieved inside the control volume, i.e., mass transfer among phases reaches equilibrium and internal forces equilibrate external ones fast enough in comparison to the interval of time between two observations of control volume properties.
%
%Pressure in each phase may differ if capillary forces are taken into account, for what extra constitutive equations relating each phase pressures would be needed.
%Capillary pressure would be function of porous medium properties given by some porous medium model, e.g., permeability and porosity; phase properties, e.g., surface tension and wettability, and the thermodynamic state.
%
%Capillary pressure is very often assumed to be function of phase saturations only and, hence, the thermodynamic state of the control volume would be completely determined by the pressure of one phase (e.g. the wetting phase) and phase saturations.
%
%Black-Oil Equations do not assume water and oil phases becoming miscible for any thermodynamic state.
%
%Capillary pressure dependence on phase saturation only is often considered as a weak assumption in contradiction with experimental observations. The hysteresis of capillary pressure exhibited when either imbibition or drainage occurs is not theoretically explained and it is accounted for by empirical relations or it is rather simply neglected.
%
%\cite{hilfer1} has proposed the inclusion of an energy equation.

@@

\subsection{Black-oil equations}

@@

\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV + \oint_{\partial \Omega} \rho_w \left( \vec{u}_w \cdot \vec{dS} \right) = \rho_w^{ST}Q_w^{ST} \, ;
\end{equation}

\item oil component:
\begin{equation}\label{eq: So1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV + \oint_{\partial \Omega} \rho_o Y_o \left(\vec{u}_o \cdot \vec{dS} \right)= \rho_o^{ST}Q_o^{ST} \, ;
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV + \oint_{\partial \Omega} \left( \rho_o Y_g \vec{u}_o + \rho_g \vec{u}_g\right) \cdot \vec{dS}= \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}
Mass concentration of oil and gas components in oil phases are denoted as $Y_o$ and $Y_g$, respectively.

@@

A common simplification in black-oil formulations is to restrict mass transport to macroscopic advection. No diffusive or dispersive term is included in the mass balance equations.

%This simplification must be looked with care when gradients of species concentration are important. For instance, when gas is being dissolved in oil phase in a reservoir under repressurization, this condition is not generally true.

@@

%Mass balance equations for each fluid phase are \citep{ertekin2001basic}:
%% Oleo
%\begin{equation}\label{eq: So1}
%\frac{\partial}{\partial t}\left(\frac{\phi(p_P) S_o}{B_o(p_o,P_b)}\right)+\nabla\cdot\left(\frac{\vec{u}_o}{B_o (p_o,P_b)}\right)=q_{os} \, ,
%\end{equation}
%% Gas
%\begin{equation}
%\begin{split}\label{eq: Sg1}
%\frac{\partial}{\partial t}\left[\phi(p_P)\left(\frac{S_g}{B_g (p_g)} +\frac{S_o R_s (p_o,P_b)}{B_o (p_o,P_b)}\right)\right] \\
%+\nabla\cdot\left(\frac{\vec{u}_o R_s (p_o,P_b)}{B_o (p_o,P_b)}+\frac{\vec{u}_g}{B_g (p_g)}\right)=q_{gs} \, ,
%\end{split}
%\end{equation}
%% Agua
%\begin{equation}\label{eq: Sw1}
%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)+\nabla\cdot\left(\frac{\vec{u}_w}{B_w(p_w)}\right)=q_{ws} \, ,
%\end{equation}
%
%where saturations must satisfy the constraint
%\begin{equation}\label{eq: S11}
%S_o+S_w+S_g=1 \, .
%\end{equation}
%
%Bubble-point pressure ($P_b$) is here defined as the minimum pressure for which gas phase is locally absent, i.e., $S_g=0$. Its local value is implicitly given by the solubility ratio ($R_s$) satisfying the equation:
%\begin{equation}\label{eq: Pb}
%R_s(p_o=P_b,P_b) = \frac{B_g(p_g)}{B_o(p_o,P_b)}\frac{S_g}{S_o} +R_s(p_o,P_b) \, .
%\end{equation}
%It must be remarked that bubble-point pressure is sometimes defined in a different manner as the bubble-point pressure for the hydrocarbon composition in oil phase only. According to this latter definition, $P_b = p_o$ for any saturated reservoir.
%
%Black-oil equations make the assumption of instantaneous mass transfer between oil and gas phases, what is only reasonable for a reservoir under depletion, when oil and gas components are homonegeously distributed in oil phase. Once constant temperature and fluid composition are assumed, diffusion and advection of species and energy become unimportante for phase equilibrium calculations.
%
%If gradients of species concentration are present within oil phase, however, as it occurs in saturated reservoirs under pressure maintenance, a combined mechanism of advection and diffusion exists and those are absent of black-oil equations. Some heuristic models have been proposed for problems where this issue is critical, as for gas injection projects. \cite{utseth1981numerical} has proposed to limit $dR_s/dt$ and neglect transport of gas component in oil phase for a simplified treatment.

Momentum balance for each phase is given by an extension of Darcy's law to multiphase flow. This extension consists in multiplying absolute (or single-phase) permeability ($\overleftrightarrow{k}$) by a scalar function called phase relative permeability ($k_{r,\alpha}$):

\begin{equation}\label{eq: exdarcy}
\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right) \, , \quad \alpha=o,w,g \, .
\end{equation}

@@

Temperature is thus a constant prescribed value and no energy balance equation is needed. Also, it follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if we restrict component number to the two pseudo-components. As temperature is assumed constant, oil phase density in saturated states may be a function of pressure alone. The same is valid for mass fractions of gas and oil components in oil phase.

@@

In order to close the system of equations, a connection among different phase pressures must be provided. Such expressions are referred to as capillary pressures and are usually employed in the following forms \citep{aziz1979petroleum}:
%capillary
\begin{equation}\label{eq: capillary_ow}
p_{c,ow}\left(\vec{k},\phi,S_w\right)=p_o-p_w \, ,
\end{equation}
\begin{equation}\label{eq: capillary_og}
p_{c,go}\left(\vec{k},\phi,S_g\right)=p_g-p_o \, .
\end{equation}

@@

%It is well known that black-oil equations possess discontinuous solutions in saturation, bubble pressure and solubility ratio fields. Discontinuities (or shocks) are introduced by initial or boundary conditions and are propagated throughout flow domain due to the hyperbolicity of equations.
%
%In initial conditions, if hydrostatic equilibrium is imposed and no capillary pressure is included, that is,
%\begin{equation}
%p_{c,ow}=p_{c,og}=0 \quad \longrightarrow \quad p_o=p_w=p_g=p\, ,
%\end{equation}
%saturation discontinuities will be present due to gravity segregation.
%
%When capillary pressure is present and it is function of phase saturations, a diffusive term is introduced in mass balance equations and shocks originated by phase displacement become smoothed:
%\begin{equation}
%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)-\nabla\cdot\left[\frac{k_{r,w} \vec{k}}{B_w(p_w)\mu_w} \left(\nabla p_o - \rho_w g\nabla z -\underbrace{\frac{d p_c}{d S_w}\nabla S_w}_{\text{water diffusion}}\right)\right]=q_{ws} \, .
%\end{equation}
%
%Discontinuities in saturation may still occur, however, if capillary pressures are functions of discontinuous porous medium properties ($phi$ and $\vec{k}$) or if phase change takes place, as diffusion of gas dissolved in oil phase is not allowed.
%Cite Bear and Herrera.

@@

We start assuming mechanical equilibrium, i.e., no bulk motion:
\begin{equation}
\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right)\equiv 0 \, .
\end{equation}
This condition implies that pressure gradients must equal gravity force per unit volume:
\begin{equation}\label{eq: p=rhogh}
\nabla p_\alpha = \rho_\alpha g\nabla z
\end{equation}
and, consequently, pressure is a function of depth only:
\begin{equation}\label{eq: p=rhogh}
\int_{p_\alpha^0}^{p_\alpha} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'},T\right)} = g\left(z - z_0\right) \, .
\end{equation}

@@

\begin{equation}
\bar{\phi} (t) = \frac{1}{\Omega}\int_{\Omega}\phi \left(\vec{x},t\right) dV \, .
\end{equation}

We proceed omitting function arguments:
\begin{equation}
\bar{S}_\alpha\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}S_\alpha \phi dV \, ,
\end{equation}

\begin{equation}
\bar{\rho_\alpha}\bar{S}_\alpha \bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha S_\alpha \phi dV \, ,
\end{equation}

\begin{equation}
\bar{\rho}_\alpha \bar{Y}_{\alpha,k}\bar{S}_\alpha \bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha Y_{\alpha,k} S_\alpha \phi dV \, .
\end{equation}

@@

\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw3}
\frac{\partial}{\partial t} \left( \bar{\rho_w}\bar{S}_w\bar{\phi} \Omega \right) = \rho_w^{ST}Q_w^{ST}
\end{equation}

\item oil component:
\begin{equation}\label{eq: So3}
\frac{\partial}{\partial t} \left( \bar{\rho_o}\bar{Y}_o\bar{S}_o\bar{\phi} \Omega \right) = \rho_o^{ST}Q_o^{ST}
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg3}
\frac{\partial}{\partial t} \left[\left( \bar{\rho}_o\bar{Y}_g \bar{S}_o + \bar{\rho}_g \bar{S}_g \right) \bar{\phi} \Omega\right] = \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}

@@

\begin{equation}\label{eq: Bw}
B_w = \frac{\rho_w^{ST}}{\bar{\rho}_w} \, ,
\end{equation}

\begin{equation}\label{eq: Bo}
B_o = \frac{\rho_o^{ST}}{\bar{Y}_o \bar{\rho}_o} \, ,
\end{equation}

\begin{equation}\label{eq: Bg}
B_g = \frac{\rho_g^{ST}}{\bar{\rho}_g} \, .
\end{equation}

@@

\begin{equation}
R_s = \frac{\rho_o^{ST} / \bar{Y}_o}{\rho_g^{ST} / \bar{Y}_g} \, ,
\end{equation}

@@

\begin{equation}\label{eq: Sw4}
\frac{\partial}{\partial t} \left(\frac{\bar{S}_w V_p}{B_w} \right) = Q_w^{ST} \, ,
\end{equation}

\begin{equation}\label{eq: So4}
\frac{\partial}{\partial t} \left(\frac{\bar{S}_o V_p}{B_o} \right) = Q_o^{ST} \, ,
\end{equation}

\begin{equation}\label{eq: Sg4}
\frac{\partial}{\partial t} \left[ V_p \left(\frac{\bar{S}_g}{B_g} + \frac{\bar{S}_o R_s}{B_o} \right) \right] = Q_g^{ST} \, ,
\end{equation}

\begin{equation}\label{eq: S14}
\bar{S}_w+\bar{S}_o+\bar{S}_g=1 \, .
\end{equation}

@@

For answering this question, we examine the two roles played by pressure in material balance equation: computation of phase density and phase equilibrium.

We shall see that for the purpose of computing the averaged phase density, we may define some reference depth and use the corresponding local density as an approximation to the averaged density without introducing any important error. Pressure increases with depth according to the hydrostatic gradient. Consequently, density also increases but by a negligible amount.

@@

For the case of an oil reservoir covered by a gas cap, the reference depth may be defined, with advantage, at the gas-oil contact. The point here is that at phase contact both gas and oil densities may be computed using the same pressure, i.e., the corresponding local value (we are neglecting capillary pressure).

@@

From Eq.~\eqref{eq: dRsdz2}, we see that the variation of pressure with depth may not be negligible for the phase equilibrium,  specially for reservoirs of light oils with large depth ranges. In such cases, using an averaged value for the solubility ratio may be worth the accuracy gain in computing dissolved gas volume.

@@

\section{Solution of material balance equation}
A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We shall start defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
\begin{equation}\label{eq: residual}
\varepsilon(p_2) = \frac{E+F_i+C-F_p}{V_{p1}} \, .
\end{equation}
We proceed showing that $\varepsilon$ is a monotonic function provided some conditions on fluid properties are observed and that, as a consequence, only one root of $\varepsilon$ exists at most. Equivalently, we may say that only one solution exists for material balance equation.

@@

\subsection{Mathematical aspects of the residual function}
